---
title: "Tiny forest project plan"
subtitle: "Submission for MOD0002205 E011"
date: "`r format(Sys.Date(), '%B %d %Y')`"
author: "SID: 2050507"
institute: "Anglia Ruskin University"
format: 
   html:
      toc: true
      toc-location: left
   docx: 
      toc: true
   pdf: 
      toc: true
editor: visual
execute: 
  message: false
  warning: false
  echo: false
  cache: true
linestretch: 1.5
citations-hover: true
bibliography: references.bib
csl: /Users/julianflowers/Zotero/styles/harvard-anglia-ruskin-university.csl
---

# A framework for evaluating the impact of Tiny Forests on biodiversity in the UK

## Abstract

The importance of ecological and ecosystem restoration is increasingly recognised... UNDECAD. Urban growth and urban restoration. Trees area vital and urban forests bring many benefits. One approach gaining popularity is the Tiny Forest (TF) movement which has its origins in Japan, but has been adopted in Europe and is rapidly develoong in the UK with more than 160 planted since 2020. TFs are largely community planted and owned and are accompanied by an extensive citizen science programme to maonitor progrress. There are claims for multiple benefits for TFs including carbon capture, human wellbeing, pollution and flood risk management and thermal comfort but increasing biodiversity is key to their success. Preliminary Citizen science biodiversity data from serial surveys of UK forests in 2022 was focused largely on butterflies, ground dwellers and pollinators. However monitoring a large and growing number of TFs creates a considerable burden. I therefore set out to establish whether more routinely available biodiversity datasets, notably the NBN Atlas could be used as a monitoring tools for TFs, and whether other satellite imagery data at high resolution (10m) could be used to to track the development of TFs.

## Introduction

Britain is one of the least biodiverse countries in Europe @whywen.

In 2021 we entered the UN Decade of Ecosystem Restoration (UNDECAD) which aims to stop ecosystem decline, in order to promote human welfare, tackle climate change and halt biodiversity collapse @undecad.

The Resolution proposing UNDECAD emphasized the importance of habitat maintenance and restoration:

> *"Emphasizing that forests, wetlands, drylands and other natural ecosystems are essential for sustainable development, poverty alleviation and improved human wellbeing"*

Growth of urbanisation - Urban restoration / nature based solutions as a means of improving wellbeing, tackling biodiversity loss, carbon

### Tiny forests

One recent approach to urban ecological restoration which is growing in popularity is the development of Tiny Forests (TF)[^1]. These are defined as small scale (100 - 200 m^2^) densely planted woodlands, planted with a variety of native trees and understorey flora.

The TF concept originated in the work of Akiro Miyawaki in Japan who applied ecological principles to forest regeneration as part of disaster recovery and mitigation[@Miyawaki1993-ji; @Miyawaki2004-di; @WOS:000078078300016]. His original ideas were adopted widely in South East Asia and now have spread west. The current approach in the UK is largely based in work in the Netherlands and is and proscribed by the Tiny Forest Handbook @noauthorundated-zc.

### Tiny forest development in UK

The first TF in the UK was planted at Witney in Oxfordshire in 2020 and more than 160 have been planted since. Current locations are shown in the map.\@ggmap

```{r location-tiny-forest, fig.cap="Location of Tiny Forests at March 2023", fig.id=TRUE}

## replace dot with tree symbol

Sys.setenv(GGMAP_GOOGLE_API_KEY= "AIzaSyBhAWbDyKX2chM2Jz0ZV22uT_8qN7YKL3Q")

needs(ggmap, tidyverse, sf, here)

tf_w <- read_csv("../data/tf_w_1.csv") |>
  mutate(year = year(when), 
         year_month = zoo::as.yearmon(when))

tf_sf <- tf_w  |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

p <- ggmap::get_googlemap(center = c(lon = -1, lat = 53.5), 
                          zoom = 6, maptype = "terrain")

ggmap(p) +
  geom_jitter(data = tf_w, aes(x = lon, y = lat, colour = factor(year))) +
  scale_colour_viridis_d(option = "plasma", direction = 1, name = "Planting year") +
  theme_void() +
  labs(title = "Planting years and locations of Tiny Forests in the UK")


```

```{css}
.mermaid {
  background-color: lightblue;
  fill: green;
  border: none;
  padding: 15px;
  border-radius: 5px;
}

```

```{mermaid, theme: forest}
%%| fig-width: 6

graph LR
    A[Tiny Forest research questions] 
    B[Flood management]
    C[Thermal comfort]
    D[Biodiversity]
    E[Carbon capture and storage]
    F[Social and wellbeing benefits]
    G[Annual carbon storage]
    H[Variation across the network]
    I[Water storage capacity]
    J[Cooling effect?]
    K[Connection to nature]
    L[Social reach]
    M[Invertebrate abundance]
    N[Variaton between TFs]
    style A fontSize:18px
    
    A --- B
    A --- C
    A --- D
    E --- A 
    F --- A
    G -.- E
    H -.- E
    B -.- I
    C -.- J
    D -.- M
    D -.- N
    K -.- F
    L -.- F 
    
    style B fill:#FFFFFF
    style C fill:#FFFFFF
    style D fill:#FFFFFF
    style E fill:#FFFFFF
    style F fill:#FFFFFF
    

```

### Existing research

Rapid preliminary literature review reveals very little research on the benefits or biodiversity impact of TFs to date. This is not surprising in a UK context given their relative novelty. Earth Watch published a snap shot of the performance of the UK Tiny Forest programme in 2022. Reporting on 359 biodiversity surveys focusing on butterflies, pollinators and ground-dwelling invertbrates, they found (excluding ants) woodlice were the predominant invertebrate, "other flies" the predominant pollinator, and small white the commonest butterfly. They found little variation between forests but there was some evidence of increase in richness of invertebrates over time @hunston.

@Ottburg2018-ls conducted periodic surveys of birds frequency, flora and and soil fauna using a standardised methodology at two TFs in Zaanstad compared with two close by forested areas in order to assess whether TFs can bring biodiversity and whether citizen science (CS) can generate biodiversity overviews. The authors found that selected pollinators and soil invertebrates were more abundant in the TFs than controls, but they didn't perform any statistical analysis and the comparabiilty of control groups wasn't stated @Ottburg2018-ls.

@cárdenas2022 outline research objectives for TFs including monitoring the seasonal and inter-forest variation in invertebrate richness and abundance, and relating it to the surrounding green infrastructure.

### Measuring biodiversity

## Aims and objectives of this research

The primary objective of this research is to

1.  Describe the current portfolio of TFs

2.  Make an initial assessment of biodiversity associated with each TF and the surrounding area

## Proposed methods

The study will largely be desk based using publicly available datasets. I propose to visit a convenience sample of TFs in my local area to help my understanding and with permission of Earth Watch talk to local TF custodians (known as Treekeepers). This will provide more context for how TFs are managed and maintained, local CS programmes, and help the utility and acceptability of this research.

### Literature review

I will undertake a rapid review of the literature on the relationship between biodiversity and urban forestry with a particular focus on Tiny Forests and similar projects. For this I propose searching Web of Science (WoS), Pubmed and Semantic Scholar with the initial search string:

> (tiny forest OR small wilderness OR pocket forest OR mini forest) AND (diversity OR biodiversity OR species diversity)

### Classifying TFs as habitat

As an initial step I propose to classify TFs according to their location, age, size and planting scheme - this will help simplify the analysis of variation and trends in biodiversity.

```{r}

here::here()

tf_w <- read_csv("../data/tf_w.csv")


  
tf_w <- tf_w |>
  mutate(stub = str_remove(url, "https://tinyforest.earthwatch.org.uk/tiny-forest-sites/tiny-forest-sites/8-tiny-forest/"), 
         tf_id = parse_number(stub), 
         tf_name = str_remove(stub, "\\d{1,3}-"), 
         age = today() - when) |>
  select(-url)



```

### Control groups

To evaluate variation and change in biodiversity I propose two controls per TF:

1.  a green space area (allotment or public park) randomly selected from the OS Green Space data within a 1km buffer of the point location of TF

2.  The nearest deciduous woodland area identified from Natural England's Priority Habitat Inventory.

For each I propose to calculate an annual time series of biodiversity metrics - species richness and Shannon Weiner diversity index - across 3 classes - Aves, Insecta and mammalia in a buffer of 500m around the centroid of each area. These 3 classes represent a subtatinal proportion of the records in teh NBN atlas and complement the CS datasets This analysls potentially gives nearly 1500 time-series (160 areas x 3 ts per area = 480, x 3 classes) so one approach might be to draw a stratified random sample of areas to reduce analytical overhead and complexity \[although power, and automation\].

### 

### Analysis plan

```{mermaid}
%%| fig-width: 6
flowchart LR
    A[[Design]] --> B[Units of analysis]
    B --> C[[Tiny forests N = 160]]
    B --> D[[Allotments ?N = 160]]
    C --> E{{1 km buffer}}
    E --> G{{Descriptive analytics}}
    E --> H{{Landuse/cover}}
    E --> I{{Summary measures over time}}
    D --> F{{1 km buffer, random sampling from 5 km buffer around TF}}
    F --> J{{Descriptive analytics}}
    F --> K{{Landuse/cover}}
    F --> L{{Summary measures over time}}
```

### Data

Proposed data sources and application

| Measure      | Source                                     | Analytical purpose                                                                  |
|--------------|--------------------------------------------|-------------------------------------------------------------------------------------|
| Biodiversity | NBN Atlas species occurrence data @nbnatla | Calculate species richness and diversity indices for buffer zones and control areas |
|              | Earth Watch data                           | Calculate diversity indices for Tiny Forests                                        |
| Tiny Forests | Earth Watch data                           | Features of TFs e.g. age, area, planting density and diversity                      |
|              | Sentinel-2 (satellite data)                | Land-cover and vegetation indices                                                   |
|              | OS                                         | Local greenspace                                                                    |
|              | Natural England                            | Local woodland                                                                      |
|              | ONS                                        | Urban rural classification, social deprivation scores                               |

: Data sources and uses

### Analytical methods

#### Descriptive analysis

-   Characteristics of TF areas
-   Summary of biodiversity
-   Trends

#### Biodiversity

-   Species richness and diversity estimates by area and time in study and control groups

### Modelling

TBD

## Project plan

```{mermaid}
%%| fig-width: 7

gantt
        dateFormat  DD-MM-YYYY
        axisFormat  %d-%m-%Y
        
        title Outline project plan
        
        section MOD003371
        Write up assignment       :active, des1, 01-04-2023,21-04-2023
        Submit ethics             :active, 21-04-2023, 1d
        Submit risk assessment    :active, 21-04-2023, 1d
        
        section MOD002979 
        Literature review         :active,  des2, 01-04-2023, 28d
        Review & assemble data    :active,  des3, 01-04-2023, 56d
        Analysis                  :         des4, after des3, 60d
        Write up                  :         des5, after des4, 120d
        Review, final edits.      :         des6, 21-10-2023, 20-12-2023
        
        section Critical tasks
        Submit MOD003371          :crit, active, 21-04-2023,24h
        Agree project outlines    :crit, done, after des1, 2d
        Submit final project  MOD002979  :milestone, 31-12-2023, 1d
        
        section Meetings
        Supervisor                : des, 13-04-2023, 1d
        Earth Watch               : des, 20-06-2023, 1d
```

## References

[^1]: Also known as mini-forests or pocket forests
