---
title: "Tiny forest project plan"
subtitle: "Submission for MOD0002205 E011"
date: "`r format(Sys.Date(), '%B %d %Y')`"
author: "SID: 2050507"
institute: "Anglia Ruskin University"
format: 
   html:
      toc: true
      toc-location: left
      toc-depth: 3
   docx: 
      toc: true
   pdf: 
      toc: true
editor: visual
execute: 
  message: false
  warning: false
  echo: false
  cache: true

citations-hover: true
bibliography: references.bib
csl: /Users/julianflowers/Zotero/styles/harvard-anglia-ruskin-university.csl
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.path = "figs/",
                      fig.align = "left")

```

# A framework for evaluating the impact of Tiny Forests on biodiversity in the UK

## Abstract

## Introduction

### The biodiversity crisis

Britain is one of the least biodiverse countries in Europe @whywen. Urban restoration

In 2021 we entered the UN Decade of Ecosystem Restoration (UNDECAD) which aims to stop ecosystem decline, in order to promote human welfare, tackle climate change and halt biodiversity collapse @undecad.

The Resolution proposing UNDECAD emphasized the importance of habitat maintenance and restoration:

> *"Emphasizing that forests, wetlands, drylands and other natural ecosystems are essential for sustainable development, poverty alleviation and improved human wellbeing"*

Growth of urbanisation - Urban restoration / nature based solutions as a means of improving wellbeing, tackling biodiversity loss, carbon

### Urban forests

### Tiny forests

One recent approach to urban ecological restoration which is growing in popularity is the development of Tiny Forests (TF)[^1]. These are defined as small scale (100 - 200 m^2^) densely planted native woodlands. There is a strong emphasis on native planting and soil preparation, community engagement and citizen science.

The TF concept originated in the work of Akiro Miyawaki in Japan who applied ecological principles to forest regeneration as part of disaster recovery and mitigation[@Miyawaki1993-ji; @Miyawaki2004-di; @WOS:000078078300016]. His original ideas were adopted widely in South East Asia and now have spread worldwide inspired by a TED talk by Shubhendu Sharma. The current approach in the UK is largely based in work in the Netherlands and is and proscribed by the Tiny Forest Handbook @noauthorundated-zc.

### Tiny forest development in UK

The first TF in the UK was planted at Witney in Oxfordshire in 2020 and more than 160 have been planted since. Current locations are shown in the map.

```{r location-tiny-forest, fig.cap="Figure \\ref{fig:fig1} Location of Tiny Forests at March 2023"}

## replace dot with tree symbol

Sys.setenv(GGMAP_GOOGLE_API_KEY= "AIzaSyBhAWbDyKX2chM2Jz0ZV22uT_8qN7YKL3Q")

needs(ggmap, tidyverse, sf, here)

tf_w <- read_csv("../data/tf_w_1.csv") |>
  mutate(year = year(when), 
         year_month = zoo::as.yearmon(when))

tf_sf <- tf_w  |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

p <- ggmap::get_googlemap(center = c(lon = -1, lat = 53.5), 
                          zoom = 6, maptype = "terrain")

ggmap(p) +
  geom_jitter(data = tf_w, aes(x = lon, y = lat, colour = factor(year))) +
  scale_colour_viridis_d(option = "plasma", direction = 1, name = "Planting year") +
  theme_void()  


```

```{css}
.mermaid {
  background-color: lightblue;
  fill: green;
  border: none;
  padding: 15px;
  border-radius: 5px;
}

```

```{mermaid}
%%| fig-width: 6
%%| fig-cap: Tiny Forest research questions

graph LR
    A[Tiny Forest research questions] 
    B[Flood management]
    C[Thermal comfort]
    D[Biodiversity]
    E[Carbon capture and storage]
    F[Social and wellbeing benefits]
    G[Annual carbon storage]
    H[Variation across the network]
    I[Water storage capacity]
    J[Cooling effect?]
    K[Connection to nature]
    L[Social reach]
    M[Invertebrate abundance]
    N[Variaton between TFs]
    style A fontSize:18px
    
    A --- B
    A --- C
    A --- D
    E --- A 
    F --- A
    G -.- E
    H -.- E
    B -.- I
    C -.- J
    D -.- M
    D -.- N
    K -.- F
    L -.- F 
    
    style B fill:#FFFFFF
    style C fill:#FFFFFF
    style D fill:#FFFFFF
    style E fill:#FFFFFF
    style F fill:#FFFFFF
    

```

### Existing research

Rapid preliminary literature review reveals very little research on the benefits or biodiversity impact of TFs to date. This is not surprising in a UK context given their relative novelty. Earth Watch published a snap shot of the performance of the UK Tiny Forest programme in 2022. Reporting on 359 biodiversity surveys focusing on butterflies, pollinators and ground-dwelling invertebrates, they found (excluding ants) woodlice were the predominant invertebrate, "other flies" the predominant pollinator, and small white the commonest butterfly. They found little variation between forests but there was some evidence of increase in richness of invertebrates over time @hunston.

@Ottburg2018-ls conducted periodic surveys of bird frequency, flora and and soil fauna using a standardised methodology at two TFs in Zaanstad compared with two close-by forested areas in order to assess whether TFs can bring biodiversity and whether citizen science (CS) can generate biodiversity overviews. The authors found that selected pollinators and soil invertebrates were more abundant in the TFs than controls, but they didn't perform any statistical analysis and the comparabiilty of control groups wasn't stated @Ottburg2018-ls.

@cárdenas2022 outline research objectives for TFs including monitoring the seasonal and inter-forest variation in invertebrate richness and abundance, and relating it to the surrounding green infrastructure.

TFs are in effect a change in land-use of amenity grassland areas but other options exist to improve biodiversity through altered land-use. For example, @Helden2004-ro estimated species richness of Hemiptera (true bugs) on urban roundabouts in Bracknell, Hampshire UK and found that native planting had greater abundance than non-native planting which in turn influenced the abundance of blue and great tits @Helden2012-wr. Allotments and community gardens (urban agriculture) have also been found to impact biodiversity although a recent systematic review found a strong North American bias, and limited taxal focus @clucas2018.

## Aims and objectives of this research

The primary objectives of this research area

1.  Characterise

2.  Make an initial assessment of biodiversity associated with each TF and the surrounding area

3.  Explore how biodiversity varies between TFs by taxa and over time

4.  Compare biodiversity estimates from directly collected data with those derived from analysis of NBN atlas data.

## Proposed methods

The study will largely be desk based using publicly available datasets. I propose to visit a convenience sample of TFs in my local area to help my understanding and with permission of Earth Watch talk to local TF custodians (known as Treekeepers). This will provide more context for how TFs are managed and maintained, local CS programmes, and help the utility and acceptability of this research.

### Literature review

I will undertake a rapid review of the literature on the relationship between biodiversity and urban forestry with a particular focus on Tiny Forests and similar land-use. For this I propose searching Web of Science (WoS), Pubmed and Semantic Scholar with the initial search string:

> (tiny forest OR small wilderness OR pocket forest OR mini forest OR allotment\* OR urban forest\* OR community garden\*) AND (diversity OR biodiversity OR species diversity OR species richness)
>
> Based on @clucas2018

### Measuring biodiversity

Measuring biodiversity is essential for monitoring the impact of any land use change but is not straightforward. It involves a number of underlying concepts: *species* *richness* which involves counting the number of different species; *abundance* which estimates the number of each species or taxa; *evenness* which measures the distribution of abundance among species or taxa; and *community composition* which takes into account the extent to which species co-occur in given locations or habitats. As a results numerous metrics and indices have been proposed to summarise and compare biodiversity by place and over time.

The simplest and most widely used index is richness (or species number) which is simply a count of recorded species, however given that all biodiversity data is based on sampling, such counts are rarely comprehensive. For many taxa, detection and species identification can be difficult; sampling effort may vary, especially where opportunistic data is used; so techniques to allow for sampling variation and under-detection and recording have been developed. I am not proposing to thorough review of available metrics in this research, but recommended richness estimation methods include Chao and Jack-knife @Helden2005-ol @Henderson2021-yr

### Classifying TFs as habitat

As an initial step I propose to classify TFs according to their location, age, size and planting scheme - this will help simplify the analysis of variation and trends in biodiversity.

```{r}

here::here()

tf_w <- read_csv("../data/tf_w.csv")


  
tf_w <- tf_w |>
  mutate(stub = str_remove(url, "https://tinyforest.earthwatch.org.uk/tiny-forest-sites/tiny-forest-sites/8-tiny-forest/"), 
         tf_id = parse_number(stub), 
         tf_name = str_remove(stub, "\\d{1,3}-"), 
         age = today() - when) |>
  select(-url)



```

### Control groups

To evaluate variation and change in biodiversity I propose two controls per TF:

1.  a green space area (allotment or public park) randomly selected from the OS Green Space data within a 1km buffer of the point location of TF

2.  The nearest deciduous woodland area identified from Natural England's Priority Habitat Inventory.

For each I propose to calculate an annual time series of biodiversity metrics - species richness and Shannon Weiner diversity index - across 3 classes - Aves, Insecta and mammalia in a buffer of 500m around the centroid of each area. These 3 classes represent a subtatinal proportion of the records in teh NBN atlas and complement the CS datasets This analysls potentially gives nearly 1500 time-series (160 areas x 3 ts per area = 480, x 3 classes) so one approach might be to draw a stratified random sample of areas to reduce analytical overhead and complexity \[although power, and automation\].

### 

### Analysis plan

```{mermaid}
%%| fig-width: 6
flowchart LR
    A[[Design]] --> B[Units of analysis]
    B --> C[[Tiny forests N = 160]]
    B --> D[[Allotments ?N = 160]]
    C --> E{{1 km buffer}}
    E --> G{{Descriptive analytics}}
    E --> H{{Landuse/cover}}
    E --> I{{Summary measures over time}}
    D --> F{{1 km buffer, random sampling from 5 km buffer around TF}}
    F --> J{{Descriptive analytics}}
    F --> K{{Landuse/cover}}
    F --> L{{Summary measures over time}}
```

### Data

The data sources I propose to use and analytical uses are outlined in the table. The National Biodiversity Network (NBN) Atlas is a large scale repository of public biodiversity data with more than 200 million records covering more than 50,000 species. Inputs are a mix of opportunisitc data from citizen science as well as more standardised datasets from national monitoring schemes, record centres and government agencies @nbnatlaa.

Earth Watch kindly provided a series of spreadsheet datasets representing the range of proposed data collections Figure

```{mermaid, earthwatch, fig.cap="Earth Watch datasets"}
%%| fig-width: 6
%%| fig-cap: Figure 1. Earth Watch datasets

graph LR
    A[Earth Watch data] 
    B[Biodiversity]
    C[Thermal comfort]
    D[Carbon capture and storage]
    
    F[Soil condition]
    G[Flood risk]
    
    I[Butterfly counts]
    
    
    
    M[Tree heights]
    
   
    P(Ground dwellers)
    Q[[Pollinators FIT count]]
    S[[Water permeability]]
    T[[Soil compaction / texture / colour]]
    U[[Soil type and chemistry]]
    V[[Soil preparation]]
    W[[Tree planting]]
    X[[Weather and perception]]
    style A fontSize:18px
      
    A --- B
    A --- C
    A --- D
    
    F --- A
    G --- A
    
    B -.- I
    B -.- P
   
    D -.- M
    
   
    
    T -.- G
    B -.- Q
    S -.- G
    U -.- F
    V -.- F
    B -.- W
    C -.- X
    
    
    
    style B fill:#FFFFFF
    style D fill:#FFFFFF
    style C fill:#FFFFFF
    style F fill:#FFFFFF
    style G fill:#FFFFFF

```

| Measure      | Source                                     | Analytical purpose                                                                  |
|------------------------|-------------------------|------------------------|
| Biodiversity | NBN Atlas species occurrence data @nbnatla | Calculate species richness and diversity indices for buffer zones and control areas |
|              | Earth Watch data                           | Calculate diversity indices for Tiny Forests                                        |
| Tiny Forests | Earth Watch data                           | Features of TFs e.g. age, area, planting density and diversity                      |
|              | Sentinel-2 (satellite data)                | Land-cover and vegetation indices                                                   |
|              | OS                                         | Local greenspace                                                                    |
|              | Natural England                            | Local woodland                                                                      |
|              | ONS                                        | Urban rural classification, social deprivation scores                               |

: Data sources and uses

### Analytical methods

#### Descriptive analysis

-   Characteristics of TF areas
-   Summary of biodiversity
-   Trends

### Example analysis: calculating biodiversity metrics for a single TF area

```{r}

source("../scripts/bdi_per_tf.R")

tychwood <- calc_bd_metrics("../large-data/nbn_tf_data.csv",   85, "Aves")

tychwood$metrics |>
  mutate_at(4:5, round, 2) |>
  knitr::kable()

```

```{r}
df <- tychwood$matrix[, -1]

specpool(tychwood$matrix[, -1])

k <- sample(nrow(df), 1)
estimateR(df[k,])

r <- poolaccum(df, minsize = 1, permutations = 9999) 

r$boot |>
  data.frame() |>
  rowwise() |>
  summarise(mean = mean(c(X1:X9999)), 
            se = sqrt(sd(c(X1:X9999), na.rm = TRUE))/ncol(r$boot)) |>
  mutate(sample = row_number(), 
         lci = mean - 1.96 * se, 
         uci = mean + 1.96 * se) |>
  ggplot(aes(as.integer(sample), mean)) +
  geom_point() +
  geom_smooth(method = "gam") +
  geom_linerange(aes(ymin = lci, ymax = uci))
  

```

## Time management and project plan

```{mermaid}
%%| fig-width: 7

gantt
        dateFormat  DD-MM-YYYY
        axisFormat  %d-%m-%Y
        
        title Outline project plan
        
        section MOD003371
        Write up assignment       :active, des1, 01-04-2023,21-04-2023
        Submit ethics             :active, 21-04-2023, 1d
        Submit risk assessment    :active, 21-04-2023, 1d
        
        section MOD002979 
        Literature review         :active,  des2, 01-04-2023, 28d
        Review & assemble data    :active,  des3, 01-04-2023, 56d
        Analysis                  :         des4, after des3, 60d
        Write up                  :         des5, after des4, 120d
        Review, final edits.      :         des6, 21-10-2023, 20-12-2023
        
        section Critical tasks
        Submit MOD003371          :crit, active, 21-04-2023,24h
        Agree project outlines    :crit, done, after des1, 2d
        Submit final project  MOD002979  :milestone, 31-12-2023, 1d
        
        section Meetings
        Supervisor                : des, 13-04-2023, 1d
        Earth Watch               : des, 20-06-2023, 1d
```

## Ethics

The ethical approval form is attached at Annex. The proposed research is largely desk-based.

## Risk management

Given the proposal is largely desk based, there will be no risks from site visits or field work.

## Potential benefits of proposed research

There is a lack of high quality research on the impact of urban restoration on biodiversity and Tiny Forests in particular. Given their increasing profile, setup costs and relative resource intensity in volunteer based monitoring and maintenance it is hoped this research will provide a framework for using routine public data to assist ongoing biodiversity assessment. It is hoped that this in turn will contribute to demonstrating the value of TFs as a form of urban ecological restoration as well as a community asset for well-being and education.

## References

[^1]: Also known as mini-forests or pocket forests
